---
title: "Time Series (Optional)"
author: "Fede Dallo"
date: "4/15/2022"
#output: html_document
output:
  #bookdown::html_document2: default
  html_notebook:
    highlight: tango
    mathjax: null
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: yes
    

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # Tom's favourite library ;)
library(lubridate) # In the Tidy universe a real shiny library to do time series analysis
library(here) # This library allows you to simplify organizing project files and directories
#library(rvest) # Not mandatory for the class, but we can use this library to get file from internet. 
```

# Global Project Options

When you start a project is good practice to use ``here::i_am("<TypeProjectName>")`` to set up the top-level project directory. The base-R command to do that is ``setwd()``, but this is fragile as it is operating system dependent as well as filesystem (absdolute and relative) dependent.

More info: <https://here.r-lib.org>

```{r, include=FALSE}
here::i_am("ARCH 249 Module 3.Rproj")
```

# Optional

## Get data files from NOAA

In the class you've been used a dataset that I've prepared. But, what if you want to perform 
your analysis? 

A great source of meteorological data is <https://www.ndbc.noaa.gov/>

### Using rvest

### Using "bash" and "wget"

```{bash}
# bash daemon is being summoned in the directory from where is called:
# in this case it is called from the file "*.Rmd" that is in the "/docs" directory

# change into "data" directory
cd ../data
pwd

for index in {11..20}

do
    FILE=ftpc1h20$index.txt.gz
    
    if [ -f "$FILE" ]; then
        echo "$FILE exists."
    else
        echo "$FILE does not exist. Downloading"
        /opt/homebrew/bin/wget https://www.ndbc.noaa.gov/data/historical/stdmet/ftpc1h20$index.txt.gz
    fi
    
done
```


### Prepare the data for analysis

```{r}
data_dir <- here("week3", "data", "tmp")
setwd(data_dir)
file_list <- Sys.glob("ftpc1h****.txt")
output_recombined_file <- file.create("recombined_ftpc1h.txt")
output_recombined_file <- file("recombined_ftpc1h.txt", "r+")
for (i in file_list) {
  input <- readLines(i)
  input <- input[-c(1,2)] # delete 1st line of the header
  #input <- input[-c(1)] # delete 2nd line of the header
  writeLines(input, output_recombined_file)
}
close(output_recombined_file)
```

