---
title: "4_forecast"
author: "Fede Dallo"
date: "4/15/2022"
#output: html_document
output:
  #bookdown::html_document2: default
  html_notebook:
    highlight: tango
    mathjax: null
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: yes
    

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # Tom's favourite library ;)
library(lubridate) # In the Tidy universe a real shiny library to do time series analysis
library(here) # This library allows you to simplify organizing project files and directories
#library(rvest) # Not mandatory for the class, but we can use this library to get file from internet. 
```

# Global options

When you start a project is good practice to set up the project directory

```{r}
here::i_am("ARCH 249 Module 3.Rproj")
```


# Optional: Using wget to get data files from NOAA

## Using rvest

## Using "bash" and "wget"

```{bash}
# bash daemon is being summoned in the directory from where is called:
# in this case it is called from the file "4_forecast.Rmd" that is in the "/docs" directory

# change into "data" directory
cd ../data
pwd

for index in {11..20}

do
    FILE=ftpc1h20$index.txt.gz
    
    if [ -f "$FILE" ]; then
        echo "$FILE exists."
    else
        echo "$FILE does not exist. Downloading"
        /opt/homebrew/bin/wget https://www.ndbc.noaa.gov/data/historical/stdmet/ftpc1h20$index.txt.gz
    fi
    
done
```


# Forecast

Reference: <https://otexts.com/fpp3/>

The predictability of an event or a quantity depends on several factors including:

1- how well we understand the factors that contribute to it;
2- how much data is available;
3- how similar the future is to the past;
4- whether the forecasts can affect the thing we are trying to forecast.

### Example 4.1

Predict tomorrow sunrise, predict tomorrow lotto numbers

"Once it has been determined what forecasts are required, it is then necessary to find or collect the data on which the forecasts will be based. The data required for forecasting may already exist. These days, a lot of data are recorded, and the forecaster’s task is often to identify where and how the required data are stored. The data may include sales records of a company, the historical demand for a product, or the unemployment rate for a geographic region. A large part of a forecaster’s time can be spent in locating and collating the available data prior to developing suitable forecasting methods."

## Qualitative forecasting

When there are no data and we have to find a better than "guesswork".

## Quantitative forecasting

When:

- numerical information about the past is available;
- it is reasonable to assume that some aspects of the past patterns will continue into the future.

### Regular and Irregular spaced time series

How to fill the gap in missing data?

## tsibble

Index variable Month, and the keys Concession and Type, would be returned **even if they were not explicitly selected** as they are required for a tsibble (to ensure each row contains a unique combination of keys and index)

## Time Plots

```{r}
meteo_01M %>%
  select(dttyDatetime, mean_T_C) %>%
  #filter(mean_T_C > -20) %>%
  ##FEDE## create daily average data:
  mutate(YEAR = lubridate::year(dttyDatetime),
         MONTH = lubridate::month(dttyDatetime),
         DAY = lubridate::day(dttyDatetime)
         ) %>%
  group_by(YEAR, MONTH, DAY) %>%
  summarise(DATE = lubridate::floor_date(x = mean(dttyDatetime, na.rm=TRUE), unit = "day"),
         TEMPERATURE = mean(mean_T_C, na.rm = T)
         ) %>%
  ungroup() %>%
  select(DATE, TEMPERATURE) %>%
  #nrow()
  ##FEDE## plot:
  ggplot(data = ., aes(x = DATE, y = TEMPERATURE)) +
  geom_line() + 
  geom_point()
```



  









