---
title: "RStudio basic tour"
author: "Federico Dallo"
date: "4/5/2022"
#output: html_document
output:
  #bookdown::html_document2: default
  html_notebook:
    highlight: tango
    mathjax: null
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: yes
    

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Welcome!

## Work remotely or on your laptop? {.tabset}

Here in Berkeley you have access to Jupyter and Rstudio: <https://datahub.berkeley.edu/>! Use this resources to try both IDEs. Limitation: we just realized you can't install external software (e.g. "eplus". We will need it in week4)

So... for this class  you will need to install R on your laptop. You can find it here: <https://cloud.r-project.org>

R do not comes with RStudio (.. make sense), which is the main IDE for R. You can install the RStudio Desktop from here: <https://www.rstudio.com/products/rstudio/>

## General interaction with R environment(s) {.tabset}

### Basic calls
```{r}
"Hello Cal!" # this is an explicit print
```

```{r}
5 # this is a number.. 
```

```{r}
is.numeric(5) # this is a number.. 
```

```{r}
typeof(5) # but what kind of number??
```

```{r}
# Can I force to have another representation?
typeof( as.integer(5) )
print( as.integer(5) ) # we need to explicitly invoke the "print" function
typeof( as.character(5) )
print( as.character(5) )
```

```{r}
for (i in 1:5) {print(i)}
```

### Excercise

## Arithmetics 
